---
title: "INFX 573 Lab: Exploring Data"
author: "Charudatta Deshpande"
date: "October 10th, 2017"
urlcolor: blue
output: 
  tufte_handout:
    highlight: tango
---

# Collaborators: N/A

# \textbf{Instructions:}

Before beginning this assignment, please ensure you have access to R and/or RStudio. You will also need to install two R packages that we will be using throughout the course.
\marginnote{\textcolor{blue}{Hint: If you encounter any errors, you might need to install other dependences, including `Rcpp` and `tibble`.}} You can install these packages in R using the following commands:

```{r Prepare, message=FALSE,eval=FALSE}
# Install packages if you don't have them
install.packages("tidyverse")
install.packages("tufte")
```

1. Download the `lab03a_titanic.rmd` file from Canvas. Open `lab03a_titanic.rmd` in RStudio (or your favorite editor) and supply your solutions to the assignment by editing `week2a_lab.Rmd`.  You will also want to download the `titanic.txt` data file, containing a data about passengers aboard the Titanic.

2. Replace the "Insert Your Name Here" text in the `author:` field with your own full name.

3. Be sure to include code chucks, figures and written explanations as necessary. Any collaborators must be listed on the top of your assignment. Any figures should be clearly labeled and appropriately referenced within the text. 

5. When you have completed the assignment and have **checked** that your code both runs in the Console and knits correctly when you click `Knit`, rename the R Markdown file to `YourLastName_YourFirstName_lab2a.Rmd`, and knit it into a PDF. Submit the compiled PDF on Canvas.

```{r Setup, message=FALSE}
# Load some helpful libraries
library(tidyverse)
```

# \textbf{Exploring Data:}

The sinking of the RMS Titanic[^titanic_ref] is a notable historical
event. The RMS Titanic was a British passenger liner that sank in the
North Atlantic Ocean in the early morning of 15 April 1912, after
colliding with an iceberg during her maiden voyage from Southampton to
New York City. Of the 2,224 passengers and crew aboard, more than
1,500 drowned, making it one of the deadliest commercial peacetime maritime disasters in modern history.

The disaster was greeted with worldwide shock and outrage at the huge loss of life and the regulatory and operational failures that had led to it. Public inquiries in Britain and the United States led to major improvements in maritime safety. One of their most important legacies was the establishment in 1914 of the International Convention for the Safety of Life at Sea (SOLAS)[^solas], which still governs maritime safety today. 

The data were originally collected by the British Board of Trade in
their investigation of the sinking. You can download these data in CSV
format from Canvas.  That there is not complete agreement among
primary sources as to the exact numbers on board, rescued, or lost.
You can find the variable definitions at [kaggle.com](https://www.kaggle.com/c/titanic/data).

## Formulate a Question:

Today, we will consider two questions in our exploration:

- Who were the Titanic passengers? What characteristics did they have?
- What passenger characteristics or other factors are associated with survival?

## Read and Inspect Data:

To begin, we need to load the Titanic dataset into R. You can do so by executing the following code.

```{r Read in Data, message=FALSE}
titanic <- read.csv("titanic.csv")
   # Note: you can read compressed files directly, no need to manually uncompress
```

Next, we want to inspect our data. We don't want to assume that are data in exactly as we expect it to be after reading it into R. It is helpful to inspect the data object, confirming to looks as expected. 

\marginnote{\textcolor{blue}{
Hint: Some helpful functions for inspecting data are: \texttt{head(), tail(), str(), nrow(), ncol(), table()}}
}

Try editing to following code chunk to look at the top and bottom of your data frame. Perform any other inspection operations you deem necessary. Do you observe anything concerning?

```{r Inspect Data, message=TRUE, tidy=TRUE}
# Convert the file into data.table format for ease of analysis and exploration. I personally use this frequently.
library(data.table)
as.data.table(titanic)
# Below code with display first few rows of the titanic CSV file. 
head(titanic)
# Below code with display last few rows of the titanic CSV file. 
tail(titanic)
```   
**Charu's observations -      
1. Not all data fields are available for all passangers, especially for the ones that did not survive.        
2. Ticket numbers are in inconsistent format. Some of the ticket numbers are same for multiple rows, which indicates one ticket was issued per family.   
3. Fare is probably not supposed to be four decimal points. It is most likely that passangers in 1912 paid only
up to cents (2 decimal places).   
4. Age is not properly formatted. A three byte integer should be enough for this analysis.    
5. One row can contain more than one cabin number.    
6. Most passangers who do not have a 'boat' assigned did not survive. 'Boat' most likely refers to a rescue boat.** 

**Continued - Data types**   
**Following data types are observed in this dataset -   
pclass - Integer   
survived - Integer (this will be later converted to character for ease of exploration)    
name - character    
sex - character   
age - Numeric (to be converted to Integer later)    
sibsp - Integer     
parch - Integer     
ticket - Character     
fare - Numeric     
cabin - character     
embarked - character     
boat - character     
body - Integer     
home.dest - character**     


\marginnote{\textcolor{red}{Note: Remember to describe your results! You should write a reponse to accompany your analysis that comments on what you find.}}

Think about the variables in this data as they are defined. Which variables might you want to re-cast to be the appropriate data type in R?

\marginnote{\textcolor{blue}{
Hint: Consider how variables are measured and how that matches available data types in R.}}

Transform the data type of varibles you identify as improperly cast.

**Charu's response -   
Age needs to be converted into integer format.     
Also, the 'survived' variable needs to be converted to character for the purpose of filling.   
Some other changes can be made, but they are not useful for the purpose of exploratory analysis.**                 

```{r Clean Data, message=TRUE}
# We will use a data.table function to change type of column age from numeric to integer.
titanic$age = as.integer(titanic$age)
titanic$survived <- as.character(titanic$survived)
```

## Trying the Easy Solution First:

First, we want to explore who the passengers aboard the Titanic were. There are many ways we might go about this. Consider for example trying to understand the ages of passengers. We can create a basic visualization to help us understand the distributions of age for Titanic passengers.

```{r, fig.cap = "Age of Passenders Aboard the Titanic", message=FALSE, warning=FALSE}
ggplot(data = titanic, aes(age)) + 
  geom_histogram(fill="blue")
```

We might go further to look at how passenger age might be related to survival.

```{r, fig.cap="Survival and Passenger Age",message=FALSE, warning=FALSE}
ggplot(data = titanic, aes(age, survived)) + 
  geom_point(size=2, alpha=0.5, color="red")
```

\marginnote{\textcolor{red}{Note: You need to add a written response here!}}

Do you like the above figure? Why or why not? Produce a new figure that you think does a better job of helping you explore the association between passenger age and survival.

**Charu's Response -**    
**The above figure (Survival and Passenger Age) does not do a good job of explaining the relationship between survival and passanger age. It is hard to understand and no conclusions can be drawn from the figure.**    

**Also the Y axis has values between 0 and 1 which are never going to be valid.**    
    
**Based on my understanding, I believe following figure will do a better job of explaining this relationship.**    

```{r, fig.cap="Survival and Passenger Age - Stacked Bar Plot",message=FALSE, warning=FALSE}
ggplot(titanic, aes(age)) + 
  geom_bar(aes(fill = survived), position="stack")
```
**Charu's Response continued -    
Based on the above figure, it is seen that generally for any age, except under age of 10, for most ages the pattern is the same. Number of people that did not survive is greater than those survived. However, under the age of 10, more children have survived. This may be due to the fact that children were rescued with priority, or as a courtesy. But for the most part, age of a passanger does not seem to have any effect on their survival**

## Do Additional Analysis

Identify one additional data feature you want to explore. Produce one
visualization that explore this feature. Describe why you think this
is interesting and what you find. 

Describe/explain what you find!

**Charu's Response -  **    
**I am interested in finding out the relationship between the effect of sex and class of travel (pclass) on the survival of a passenger. It was implied by some critics that the wealthy passengers travelling in first class had an unfair advantage of having better access to emergency supplies like life jackets, rescue boats etc. While those travelling in third class, mostly working class passengers, were claimed to have very limited access to these supplies, and it was implied that the working class passangers had a less survival rate compared to wealthy passangers. I would like to explore if this is true. Also, it is claimed that women were given priority for rescue, so their survival rate was better than men. We will also analyze if this is true.**    
    
**We will plot the jitter plot. It is suitable since both pclass and sex are discrete variables.**      


```{r, fig.cap = "Effect of Passanger Class and Sex on Survival", message=TRUE, warning=FALSE}
ggplot(titanic, aes(sex, pclass, color=survived)) + geom_jitter(size=0.6)
```
**Charu's response continued - **  
**All of the claimed theories appear to be true.**   
**1. In all passanger classes, female survival rate is better than males for a specific passenger class. This indicates that women were rescued with priority.**   
**2. Chances of survival decrease as the passenger class decreases from first to third. This does indicate that first class wealthy passengers had better access to rescue services than working class passengers in third class.**  
**3. Males in third class constitute the biggest share of the victims (not survived).**   
**4. Almost all women in first class survived. The chances of survival are extremely high in this group.**   

## What Next?

\marginnote{\textcolor{red}{Note: You need to add a written response here!}}

Consider the exploratory analysis you completed in the lab exercise. What would you do next?

**Charu's Response - **    
**Based on this exploratory analysis, further analysis can be done in following ways -    
1. Is there a relationship between 'boats' and 'survived'? Boats variable is not explained on Kaggle. However, it may indicate the use of a rescue boat if the passenger used it. Did getting a boat help the passenger survive? Did anyone not survive after getting a boat?     
2. Is there a relationship between getting a boat and class of travel? Did wealthier passengers have better access to boats?    
3. Some outside data is needed for this specific analysis. How were the rooms structured for different classes? Did the location of first and second class allow easy access to decks and rescue boats? Did third class passengers have poor access to these because of their locations?    
4. Did the first class and second class passengers actually pay more fare than third class?      
5. For those travelling with families, did all family members survive,or did all family members die? Can a relationship be found where only children survived and parents died?     
5. Use machine learning methods to calculate best possibilty of survival using age, sex, boat, class and travelling companions. Which group fares the best and why?** 

[^titanic_ref]: https://en.wikipedia.org/wiki/RMS_Titanic
[^solas]: https://en.wikipedia.org/wiki/International_Convention_for_the_Safety_of_Life_at_Sea
